---
title: "Clustering genes based on HPO associations"
output: html_notebook
---

The Human Phenotype Ontology (HPO) team have created a standardised vocabulary of phenotypic abnormalities and associated them to genes. These associations are in the file `ALL_SOURCES_ALL_FREQUENCIES_phenotype_to_genes.txt.gz`, which I have saved in the `data` folder of this repository.

```{r}
setwd('~/github/human_phenotype_ontology/gene')
pg <- read.table('../data/ALL_SOURCES_ALL_FREQUENCIES_phenotype_to_genes.txt.gz', header = FALSE, stringsAsFactors = FALSE, quote='', sep="\t", comment='#')
names(pg) <- c('hpo', 'hpo_def', 'entrez', 'gene')
head(pg)

length(unique(pg$gene))
```

On average, how many HPO terms are associated with genes?

```{r}
library(dplyr)
pg %>%
  select(hpo, gene) %>%
  group_by(gene) %>%
  summarise(n = n()) %>%
  summarise(total = n(), mean = mean(n), median = median(n))
```

Create a function that retrieves the associated HPO terms for a gene symbol.

```{r}
gene_to_hpo <- function(x){
  r <- pg %>% filter(gene == x) %>% select(hpo)
  return(r$hpo)
}
gene_to_hpo('SLC39A5')
```

I will use the Jaccard index to measure similarity.

```{r}
jaccard_index <- function(x, y){
  i <- length(intersect(x, y))
  d <- i/length(union(x,y))
  return(d)
}
jaccard_index(gene_to_hpo('CREBBP'), gene_to_hpo('EP300'))
```

How many pairwise comparisons would there be if I wanted to pre-calculate all the Jaccard indexes?

```{r}
choose(length(unique(pg$gene)), 2)
```

We'll use the `combn()` function to create a matrix of values containing all pairwise combinations.

```{r}
my_gene <- unique(pg$gene)
my_jaccard_indexes <- apply(expand.grid(1:length(my_gene), 1:length(my_gene)),
                            1,
                            function(x) jaccard_index(gene_to_hpo(my_gene[x[1]]), gene_to_hpo(my_gene[x[2]]))
                            )

my_jaccard_matrix <- matrix(my_jaccard_indexes, nrow = length(my_gene))

row.names(my_jaccard_matrix) <- my_gene
colnames(my_jaccard_matrix) <- my_gene
save(my_jaccard_matrix, file = 'my_jaccard_matrix_all.Robj')
```

Let's perform some sanity checks.

```{r}
jaccard_index(gene_to_hpo('CREBBP'), gene_to_hpo('EP300'))

my_jaccard_matrix[grep(pattern = 'EP300', x = my_gene), grep(pattern = 'CREBBP', x = my_gene)]
my_jaccard_matrix[grep(pattern = 'CREBBP', x = my_gene), grep(pattern = 'EP300', x = my_gene)]
```

