---
title: "Clustering OMIM disorders"
output: html_notebook
---

The Human Phenotype Ontology (HPO) team have created a standardised vocabulary of phenotypic abnormalities and associated them to OMIM disorders. These associations are in the file `phenotype_annotation.tab.gz`, which I have saved in this repository.

```{r}
pa <- read.table('../data/phenotype_annotation.tab.gz', header = FALSE, stringsAsFactors = FALSE, quote='', sep="\t", comment='')
names(pa) <- c('DB', 'DB_Object_ID', 'DB_Name', 'Qualifier', 'HPO', 'DB_Reference', 'Evidence_code', 'Onset_modifier', 'Frequency_modifier', 'With', 'Aspect', 'Synonym', 'Date', 'Assigned_by')
head(pa)
```

HPO terms are also associated to Orphanet and DECIPHER disorders but I will be using only OMIM disorders for now.

```{r}
unique(pa$DB)
```

I'll use the `dplyr` package to make it easier to work with data frames. I'll use the `filter()` function to create an OMIM data frame.

```{r}
library(dplyr)
omim <- filter(pa, DB == 'OMIM')
omim <- tbl_df(omim)
head(omim)
```

On average, how many HPO terms are associated with OMIM disorders?

```{r}
omim %>%
  select(DB_Object_ID) %>%
  group_by(DB_Object_ID) %>%
  summarise(n = n()) %>%
  summarise(total = n(), mean = mean(n), median = median(n))
```

Create a function that retrieves the associated HPO terms for an OMIM ID.

```{r}
omim_to_hpo <- function(x){
  r <- omim %>% filter(DB_Object_ID == x) %>% select(HPO)
  return(r$HPO)
}
omim_to_hpo(100050)
```

Now we're getting to the main aim of this work: I want to find out how similar a pair of OMIM disorders are based on their associated HPO terms. I will use the Jaccard index to measure similarity.

```{r}
jaccard_index <- function(x, y){
  i <- length(intersect(x, y))
  d <- i/length(union(x,y))
  return(d)
}
```

Four disorders I'm interested in are KBG syndrome (148050), Aarskog-Scott syndrome (305400), Kleefstra syndrome (610253), and Filippi syndrome (272440).

```{r}
kbgs      <- omim_to_hpo(148050)
ass       <- omim_to_hpo(305400)
kleefstra <- omim_to_hpo(610253)
filippi   <- omim_to_hpo(272440)

kbgs
```

Calculate the Jaccard indexes between the four disorders.

```{r}
jaccard_index(kbgs, ass)
jaccard_index(kbgs, kleefstra)
jaccard_index(kbgs, filippi)
jaccard_index(ass, kleefstra)
jaccard_index(ass, filippi)
jaccard_index(kleefstra, filippi)
```

The Jaccard index between KBG syndrome and Aarskog-Scott syndrome was the highest between all the pairwise comparisons. Let's compare KBG syndrome to all the other 6,946 OMIM disorders, to see where Aarskog-Scott syndrome ranks.

```{r}
my_omim <- 148050
all_omim_id <- omim %>% select(DB_Object_ID) %>% unique()
all_omim_id <- all_omim_id$DB_Object_ID
all_omim_id <- all_omim_id[!grepl(my_omim, all_omim_id)]
length(all_omim_id)
one_to_all <- sapply(X = all_omim_id, function(x) jaccard_index(omim_to_hpo(x), omim_to_hpo(my_omim)))
head(one_to_all)
```

We have the scores but we don't know which OMIM disorders they correspond to. Let's fix that by creating a function to obtain the names of an OMIM disorder.

```{r rows.print=20}
get_omim_name <- function(x){
  my_name <- pa %>% filter(DB == 'OMIM', DB_Object_ID == x) %>% select(DB_Name) %>% summarise(name = unique(DB_Name))
  my_name <- strsplit(x = unlist(my_name), split = ";")[[1]][1]
  return(my_name)
}

all_omim_name <- sapply(X = all_omim_id, FUN = get_omim_name)
one_to_all_df <- data.frame(omim = all_omim_id, jaccard = one_to_all, name = unlist(all_omim_name))
one_to_all_df %>% arrange(desc(jaccard)) %>% select(name, jaccard) %>% head(n = 20)
```

How many pairwise comparisons would there be if I wanted to pre-calculate all the Jaccard indexes?

```{r}
choose(length(unique(omim$DB_Object_ID)), 2)
```

Let's test it on a smaller data set first instead of computing 24 million Jaccard indexes.

```{r}
test_set <- omim %>% select(DB_Object_ID) %>% group_by(DB_Object_ID) %>% summarise(n = n()) %>% filter(n > 30, n < 61)
choose(nrow(test_set), 2)
```

```{r}
test_omim_id <- test_set$DB_Object_ID
test_omim_hpo <- sapply(X = test_omim_id, FUN = omim_to_hpo)
names(test_omim_hpo) <- test_omim_id
head(test_omim_hpo, 2)
```

Now to calculate all the pairwise Jaccard indexes. We'll use the `combn()` function to create a matrix of values containing all pairwise combinations.

```{r}
my_jaccard_indexes <- apply(combn(1:length(test_omim_id), 2), 2, function(x) jaccard_index(test_omim_hpo[[x[1]]], test_omim_hpo[[x[2]]]))
length(my_jaccard_indexes)
head(my_jaccard_indexes)
summary(my_jaccard_indexes)
table(my_jaccard_indexes > jaccard_index(kbgs, ass))
```

Let's store the Jaccard indexes in a matrix instead.

```{r}
my_jaccard_matrix <- diag(length(test_omim_id))
my_jaccard_matrix[lower.tri(my_jaccard_matrix)] <- my_jaccard_indexes
row.names(my_jaccard_matrix) <- test_omim_id
colnames(my_jaccard_matrix) <- test_omim_id

my_jaccard_matrix[1:6, 1:6]
```

Let's perform some sanity checks.

```{r}
x <- test_omim_hpo$`148050`
y <- test_omim_hpo$`305400`
jaccard_index(x, y)
jaccard_index(kbgs, ass)

my_jaccard_matrix[grep(pattern = 305400, x = test_omim_id), grep(pattern = 148050, x = test_omim_id)]
```

Now that we have a matrix of Jaccard indexes, we can convert this into a distance matrix and perform hierarchical clustering.

```{r, fig.width = 10}
my_jaccard_distance <- as.dist(1 - my_jaccard_matrix)
hc <- hclust(my_jaccard_distance, method = 'complete')
plot(hc, labels = FALSE)
```

Define some clusters by cutting the tree.

```{r}
table(my_jaccard_distance < 0.80)
my_cluster <- cutree(hc, h = 0.80)
table(my_cluster)[table(my_cluster)>1]
```

Cluster 76 has five OMIM disorders; what are they?

```{r}
cluster_76 <- test_omim_id[my_cluster == 76]
lapply(X = cluster_76, FUN = get_omim_name)
```

Let's cut the tree at 0.90 instead.

```{r}
table(my_jaccard_distance < 0.9)
my_cluster_h90 <- cutree(hc, h = 0.9)
table(my_cluster_h90)[table(my_cluster_h90)>1]
```

Check out cluster 60.

```{r}
cluster_60 <- test_omim_id[my_cluster_h90 == 60]
lapply(X = cluster_60, FUN = get_omim_name)
```

Notice that OMIM:177980 wasn't even in the top 20, when I compared KBG syndrome to all the other OMIM disorders?

```{r rows.print=20}
one_to_all_df %>% arrange(desc(jaccard)) %>% select(name, jaccard) %>% head(n = 20)
```

This is due to the way hierarchical clustering works. For example, OMIM:607721 is ranked two in the list above but isn't in the cluster when I cut the tree at 0.90. It is likely that OMIM:607721 matches other disorders better than it matches KBG syndrome, and is clustered with them.

```{r}
my_cluster_h90[grep(pattern = 607721, test_omim_id)]
cluster_268 <- test_omim_id[my_cluster_h90 == 268]
lapply(X = cluster_268, FUN = get_omim_name)
```

Calculate the pairwise Jaccard indexes.

```{r}
cdd <- omim_to_hpo(601853)
cs  <- omim_to_hpo(605627)
lah <- omim_to_hpo(607721)

jaccard_index(lah, cs)
jaccard_index(lah, cdd)
jaccard_index(cdd, cs)
```



